ODGI: scalable tools for pangenome graphs

### 2021-05-31 manuscript/main.tex
use \odgi\

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, ODGI importantly differs in
    its design to support highly complex pangenome graphs. The majority of the ODGI tools are implemented in an
    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
    implementations would require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, ODGI tools can build the internal graph representation (the ODGI
    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
    efficiently scaling up to hundreds of large vertebrate genomes.

    ODGI is entirely focused on "first-order" operations on the graph
    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
    graphs from raw sequences, map reads, or call variants. It
    supports interrogation of the graph for basic dimensional
    properties, positional and path-related queries, and visualization
    in one-dimension (1D) and two-dimensions (2D). ODGI also supports
    operations on the graph such as subset, subdivide, break, combine,
    normalize, or order its components, nodes, and paths. Most of the
    tools are designed to be applied together, piping the output of
    one tool into the next, thereby preventing the creation of
    intermediate files, and reducing the number of IO operations.

    ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
    applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
    implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
    we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
    The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
    node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
    how the embedded paths walk across the node. Each step records the path identifier and the previous and next
    path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
    \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
    rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
    graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
    stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node structure includes a
    byte-width mutex lock.

@@

    \topic{odgi stats} users can retrieve metrics describing the graph properties, such as the number of
    nodes, edges, paths, and graph length. ODGI also offers more advanced tools for interrogating the graphs.
    \texttt{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
    gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to
    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
    detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
    process as well.

    ODGI allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
    removes the cycles, reducing the complexity of the graph topology. \texttt{odgi groom} removes spurious inverting
    links by exploring the graph from the orientation supported by most paths. To enable efficient sequence alignment
    against the graph, long nodes can be divided into shorter nodes at a maximum requested size using
    \texttt{odgi chop}. Partial order alignment, which consists of aligning sequences against a directed acyclic graph
    (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp
    long nodes; \texttt{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor
    the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

    Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural
    variants would join the components into bigger ones). \texttt{odgi explode} separates the connected components in
    different ODGI format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same ODGI format
    file, preventing node ID collisions. \texttt{odgi extract} allows extracting specific subregions of the graph as
    defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only
    with the extracted region.

@@

        \begin{minipage}[b]{0.48\textwidth}
            \includegraphics[width=\linewidth]{fig/mhc-pangenome.eps}
            \caption{(b) Visualizing the Major Histocompatibility Complex
                (MHC) \textit{locus} pangenome. A) odgi viz 1D visualization
                of 8 haploid, phased human genome assemblies plus the chm13
                and GRCh38 reference genomes. Graph nodes’ are arranged
                from left to right. The colored bars represent the
                linearized renderings of the embedded paths versus the nodes
                in a binary matrix. The black lines under the paths are the
                links, which represent the graph topology. B)
                Bandage~\citep{26099265} representation of the graph
                topology of a consensus MHC pangenome graph representing
                only variations bigger than 100 bps. Gene labels were
                applied exploiting \texttt{odgi position} to liftover gene
                coordinates from the full graph into the consensus graph.  }

@@

    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using ODGI and VG. The
    example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome
    assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data
    Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes. All commands are provided
    in the Supplementary Material.

    \begin{table}[!t]
        \processtable{Centromeres extraction comparison using ODGI and VG
        \label{tab:02}} {
            \begin{tabular}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB) \\
                \midrule
                row1               & row1 & row1            & row1        & row1             \\
                row2               & row2 & row2            & row2        & row2             \\
                row3               & row3 & row3            & row3        & row3             \\
                row4               & row4 & row4            & row4        & row4             \\
                \botrule
            \end{tabular}}

@@

    To work with ODGI and VG it is first necessary to convert the graph from GFA to ODGI and VG format, respectively.
    Thanks to the ODGI format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
    \texttt{vg convert} in the same task. For the same reason, \texttt{odgi extract} results 20 times faster than
    \texttt{vg chunk} in extracting a portion from the full graph.

    \section{Discussion}
    We implemented a tool suite, ODGI, which is simple to work with to analyse and manipulate pangenome graphs at a
    population scale. ODGI's path implementation allows working with graphs of any complexity, without particular
    limitations. We have demonstrated that ODGI tools efficiently perform with pangenome graphs embedding full human
    chromosomes. ODGI is not only a tool suite but also a library, already used in pangenome graphs building
    (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
    framework where to implement algorithms working on variation graphs. Therefore, ODGI offers an efficient and
    extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase
    scale that characterize the pangenomic era.

>>>>

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
    implementations would require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
    efficiently scaling up to hundreds of large vertebrate genomes.

    \odgi\ is entirely focused on "first-order" operations on the graph
    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
    graphs from raw sequences, map reads, or call variants. It
    supports interrogation of the graph for basic dimensional
    properties, positional and path-related queries, and visualization
    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
    operations on the graph such as subset, subdivide, break, combine,
    normalize, or order its components, nodes, and paths. Most of the
    tools are designed to be applied together, piping the output of
    one tool into the next, thereby preventing the creation of
    intermediate files, and reducing the number of IO operations.

    \odgi\ is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
    applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
    implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
    we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
    The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
    node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
    how the embedded paths walk across the node. Each step records the path identifier and the previous and next
    path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
    \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
    rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
    graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
    stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node structure includes a
    byte-width mutex lock.

@@

    \topic{odgi stats} users can retrieve metrics describing the graph properties, such as the number of
    nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
    \texttt{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
    gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to
    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
    detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
    process as well.

    \odgi\ allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
    removes the cycles, reducing the complexity of the graph topology. \texttt{odgi groom} removes spurious inverting
    links by exploring the graph from the orientation supported by most paths. To enable efficient sequence alignment
    against the graph, long nodes can be divided into shorter nodes at a maximum requested size using
    \texttt{odgi chop}. Partial order alignment, which consists of aligning sequences against a directed acyclic graph
    (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp
    long nodes; \texttt{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor
    the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

    Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural
    variants would join the components into bigger ones). \texttt{odgi explode} separates the connected components in
    different \odgi\ format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same \odgi\ format
    file, preventing node ID collisions. \texttt{odgi extract} allows extracting specific subregions of the graph as
    defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only
    with the extracted region.

@@

        \begin{minipage}[b]{0.48\textwidth}
            \includegraphics[width=\linewidth]{fig/mhc-pangenome.eps}
            \caption{(b) Visualizing the Major Histocompatibility Complex
                (MHC) \textit{locus} pangenome. A) \texttt{odgi viz} 1D visualization
                of 8 haploid, phased human genome assemblies plus the chm13
                and GRCh38 reference genomes. Graph nodes’ are arranged
                from left to right. The colored bars represent the
                linearized renderings of the embedded paths versus the nodes
                in a binary matrix. The black lines under the paths are the
                links, which represent the graph topology. B)
                Bandage~\citep{26099265} representation of the graph
                topology of a consensus MHC pangenome graph representing
                only variations bigger than 100 bps. Gene labels were
                applied exploiting \texttt{odgi position} to liftover gene
                coordinates from the full graph into the consensus graph.  }

@@

    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG. The
    example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome
    assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data
    Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes. All commands are provided
    in the Supplementary Material.

    \begin{table}[!t]
        \processtable{Centromeres extraction comparison using \odgi\ and VG
        \label{tab:02}} {
            \begin{tabular}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB) \\
                \midrule
                row1               & row1 & row1            & row1        & row1             \\
                row2               & row2 & row2            & row2        & row2             \\
                row3               & row3 & row3            & row3        & row3             \\
                row4               & row4 & row4            & row4        & row4             \\
                \botrule
            \end{tabular}}

@@

    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
    Thanks to the \odgi\ format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
    \texttt{vg convert} in the same task. For the same reason, \texttt{odgi extract} results 20 times faster than
    \texttt{vg chunk} in extracting a portion from the full graph.

    \section{Discussion}
    We implemented a tool suite, \odgi\, which is simple to work with to analyse and manipulate pangenome graphs at a
    population scale. \odgi\'s path implementation allows working with graphs of any complexity, without particular
    limitations. We have demonstrated that \odgi\ tools efficiently perform with pangenome graphs embedding full human
    chromosomes. \odgi\ is not only a tool suite but also a library, already used in pangenome graphs building
    (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
    framework where to implement algorithms working on variation graphs. Therefore, \odgi\ offers an efficient and
    extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase
    scale that characterize the pangenomic era.
